---
title: "Homework 3 starting with R"
author: "Nikita"
date: "2022-09-01"
output: html_document
---
# Instructions: #
For this assignment, you need to answer a couple questions with
code and then take a screenshot of your working environment.

Submit the solutions including the URL to the screenshot in a doc/pdf to
Brightspace.

## 1) ## 
**Use R to figure out how many elements in the vector below are greater than 2 and then tell me what their sum (of the larger than 2 elements) is.**

```{r, defining vector and cleaning up, eval=TRUE}
#defining a vector with cleaned up data
rooms <- c(1, 2, 4, 5, 1, 3, 1, NA, 3, 1, 3, 2, 1, NA, 1, 8, 3, 1, 4, NA, 1, 3, 1, 2, 1, 7, 1, 9, 3, NA)
rooms[!is.na(rooms)]
rooms_clean <- rooms[!is.na(rooms)]
 
```

```{r, defining and printing new vector with elemtents greater than 2, eval=TRUE}
#defining and printing the new vector containing only valuers greater than 2
rooms_over_2 <- rooms_clean[rooms_clean > 2]
rooms_over_2
```

```{r, sum of elements greater than 2, eval=TRUE}
#sum of the elements larger than two
length(rooms_over_2)
```

## 2) ##
**What type of data is in the 'rooms' vector?**  
*The rooms vector is an atomic numeric vector: it contains integers, although the NA are characters, and by that had to be omitted*

## 3) ##
**Submit the following image to Github:**  
Inside your R Project (.Rproj) install the 'tidyverse' package and use the download.file() and read_csv() function to read the SAFI_clean.csv dataset into your R project as 'interviews' digitalobject (see instructions in https://datacarpentry.org/r-socialsci/setup.html and 'Starting with Data' section).
Take a screenshot of your RStudio interface showing
* a) the line of code you used to create the object *
* b) the 'interviews' object in the Environment * 
* c) the file structure of your R project in the bottom right "Files" pane. '

```{r load tidyverse, eval=TRUE, echo=TRUE}
library(tidyverse)
```


**a) the line of code you used to create the object,**
```{r, creating an interviews object from file, eval=TRUE, echo=TRUE}
interviews <- read_csv("../../data/SAFI_clean.csv", na = "NULL")
```

**b) the 'interviews' object in the Environment**
```{r the interview object in the envronment, eval=TRUE, echo=TRUE}
interviews
```
*to view the object in a separate tab in the workspace*
```{r, in order to view the interview in a different tab, eval=TRUE, echo=TRUE}
view(interviews)
```

**c) the file structure of your R project in the bottom right "Files" pane.**
Save the screenshot as an image and put it in your AUID_lastname_firstname repository inside our Github organisation (github.com/Digital-Methods-HASS) or equivalent. Place here the URL leading to the screenshot in your repository.
*please note that the 'data' folder is inside the HOMEWORK folder, due to the Danish_monarchs.csv file being used both in this assignment and the next*
https://github.com/Digital-Methods-HASS/201906799_Myrting_Nikita/blob/dc730e8cf8b6fd58318ab678daa81fd669dde536/Homework_3_Rstudio_working-environment.png

##4) Challenge: If you managed to create your own Danish king dataset, use it. If not, you the one attached to this assignment (it might need to be cleaned up a bit). Load the dataset into R as a tibble. Calculate the mean() and median() duration of rule over time and find the three monarchs ruling the longest. How many days did they rule (accounting for transition year?)

```{r loading packages, eval=TRUE, echo=FALSE}
library(tibble)
library(dplyr)
```
```{r reading the dataset and viewing it as a tiblle, eval=TRUE, echo=TRUE}
danish_monarchs <- read_csv2("../../data/Danish_monarchs.csv")
danish_monarchs <- na.omit(danish_monarchs)
as_tibble(danish_monarchs)
```

```{r calculating the mean of years_reign, eval=TRUE, include=TRUE}
mean(danish_monarchs$years_reign)
```

```{r calculating the median of years_reign, eval=TRUE, include=TRUE}
median(danish_monarchs$years_reign)
```


```{r ordering the data after years_reign, from highest to lowest, eval=TRUE, echo=TRUE}
top_monarchs <- danish_monarchs [order(danish_monarchs$years_reign, decreasing = TRUE),]
#I am stuck, I have created a new data_frame, and the data has been ordered fro mhighest to lowest, but I do not know how to make it ONLY contain these three

```

